Predicting Ambient Air Pollution (PM2.5) Across the Contiguous U.S.

### 2023-04-24 Draft.Rmd
uuu

The Extreme Gradient Boosting (XGB) model performs by training weak learners, models that have low prediction accuracy, sequentially to become a single strong learner, a model that has strong prediction accuracy. We preprocess the training set to make the predictive modeling process more accurate and smoother by prepping (estimates the quantities for 'train_scaled') and baking (assigns prepped data to an object) the data. The aforementioned top 10 predictors were used in this preprocessing recipe. Moreover, the categorical variables were factored and features with no variance were removed before this data was baked. Then, this processed data is assigned to 'folds3' and resampled on 5 subsets of the training set. Using 'fold3', we tune the hyperparameters to make the modeling process more efficient and perform grid specification to determine which hyperparameter values have high prediction accuracy (i.e. low prediction error). Next, we isolated the best hyperparameter values, so they can be used for the final boosting model. Lastly, we fitted the model and evaluated its prediction performance. Its RMSE score was 0.23, indicating the XG Boosting is an adequate model. This website was used: [RBloggers](https://www.r-bloggers.com/2020/05/using-xgboost-with-tidymodels/)

@@

As seen on the table and visual, LR has the highest RMSE and lowest R-squared estimate. KNN has the second highest RMSE with a moderate R-squared estimate. And XGB has the lowest RMSE of 0.23 and the highest R-squared estimate of 0.99. Random Forest model has the both the second lowest RMSE and second highest R-squared estimate. Since we are choosing the best model by the best RMSE, which is the lowest one, the Xgboost model is thus the best and final model based on our procedure. After fitting the Xgboost model to the testing data, we find that its RMSE and R-squared is 1.38 and 0.68, respectively. Since we are trying to predict annual average ambient air pollution of a given monitor, then these statistics mean that the amount of error in our model's predictions is about 1.38 $\mu$g/m$^3$, and that the features we used in our model are able to explain 68% of the variance in the data, which means this model does not fit/predict the data well.

@@

@@

```{r Q3-2}
set.seed(123)
# XGB model w/only cmaq
# Preprocessing recipe
rec6 <- 
  recipes::recipe(value ~ CMAQ, data = training(dat_split)) %>%
  # Convert categorical variables to factors
  recipes::step_string2factor(all_nominal()) %>%
  # Combine low frequency factor levels
  recipes::step_other(all_nominal(), threshold = 0.01) %>%
  # Remove no variance predictors which provide no predictive information 
  recipes::step_nzv(all_nominal()) %>%
  prep()

folds6 <- recipes::bake(rec6, new_data = training(dat_split)) %>%  
  rsample::vfold_cv(v = 5)

xgboost_wf <- workflows::workflow() %>%
  add_model(xgboost_model) %>% 
  add_formula(value ~ CMAQ)

# Hyperparameter tuning
# Takes a long time to run, 5-8 min
xgboost_tuned <- tune::tune_grid(object = xgboost_wf, resamples = folds6,
  grid = xgboost_grid, metrics = yardstick::metric_set(rmse, rsq),
  control = tune::control_grid(verbose = TRUE))

# Eval. model performance on train data
test_processed <- bake(rec6,  new_data = testing(dat_split))
train_processed <- bake(rec6,  new_data = training(dat_split))

test_prediction = xgboost_model_final |> 
  parsnip::fit(formula = value ~ CMAQ, data=train_processed) |> 
  predict(new_data=test_processed) |> 
  bind_cols(testing(dat_split))

res6 = test_prediction |> 
  yardstick::metrics(value, .pred) |> 
  mutate(.estimate=format(round(.estimate, 2), big.mark=","))
res6
```

```{r Q3-3}
set.seed(123)
# XGB model w/only aod
# Preprocessing recipe
rec7 <- 
  recipes::recipe(value ~ aod, data = training(dat_split)) %>%
  # Convert categorical variables to factors
  recipes::step_string2factor(all_nominal()) %>%
  # Combine low frequency factor levels
  recipes::step_other(all_nominal(), threshold = 0.01) %>%
  # Remove no variance predictors which provide no predictive information 
  recipes::step_nzv(all_nominal()) %>%
  prep()

folds7 <- recipes::bake(rec7, new_data = training(dat_split)) %>%  
  rsample::vfold_cv(v = 5)

xgboost_wf <- workflows::workflow() %>%
  add_model(xgboost_model) %>% 
  add_formula(value ~ aod)

# Hyperparameter tuning
# Takes a long time to run, 5-8 min
xgboost_tuned <- tune::tune_grid(object = xgboost_wf, resamples = folds7,
  grid = xgboost_grid, metrics = yardstick::metric_set(rmse, rsq),
  control = tune::control_grid(verbose = TRUE))

# Eval. model performance on train data
test_processed <- bake(rec7,  new_data = testing(dat_split))
train_processed <- bake(rec7,  new_data = training(dat_split))

test_prediction = xgboost_model_final |> 
  parsnip::fit(formula = value ~ aod, data=train_processed) |> 
  predict(new_data=test_processed) |> 
  bind_cols(testing(dat_split))

res7 = test_prediction |> 
  yardstick::metrics(value, .pred) |> 
  mutate(.estimate=format(round(.estimate, 2), big.mark=","))
res7
```

```{r Q3-4}
set.seed(123)
# XGB model w/o cmaq and aod
# Preprocessing recipe
rec8 <- 
  recipes::recipe(value ~ lat+county_area+lon, data = training(dat_split)) %>%
  # Convert categorical variables to factors
  recipes::step_string2factor(all_nominal()) %>%
  # Combine low frequency factor levels
  recipes::step_other(all_nominal(), threshold = 0.01) %>%
  # Remove no variance predictors which provide no predictive information 
  recipes::step_nzv(all_nominal()) %>%
  prep()

@@

# Eval. model performance on train data
test_processed <- bake(rec8,  new_data = testing(dat_split))

@@

3. There is interest in developing more cost-effect approaches to monitoring air pollution on the ground. Two candidates for replacing the use of ground-based monitors are numerical models like CMAQ and satellite-based observations such as AOD. How well do CMAQ and AOD predict ground-level concentrations of PM2.5? How does the prediction performance of your model change when CMAQ or AOD are included (or not included) in the model?

CMAQ and aod do not predict

**4. The dataset here did not include data from Alaska or Hawaii. Do you think your model will perform well or not in those two states? Explain your reasoning.**
EDIT THIS
Given that the top 10 features were 'CMAQ,' 'lat,' 'county_area,' 'lon,' 'aod,' 'popdens_county,' 'zcta,' 'log_nei_2008_pm25_sum_10000,' 'fips,' and 'log_nei_2008_pm10_sum_10000,' we think that XG Boost model will perform well since the original dataset included states of varying sizes with different climates and pollution levels. Moreover, its RMSE score is 0.62 and R-squared is 0.94, indicating that the prediction performance of this model is good in relation to the other models. Since Hawaii and Alaska are almost polar opposites with the former being geographically smaller and having hotter climate, we suspect that the model will be able to predict PM2.5 concentration levels in Hawaii and Alaska if they were included in this dataset since the diversity but also inter-relatedness of the features in this dataset aids the model in predicting ambient air pollution in differing environments.

@@

@@

>>>>

The Extreme Gradient Boosting (XGB) model performs by training weak learners, models that have low prediction accuracy, sequentially to become a single strong learner, a model that has strong prediction accuracy. We preprocess the training set to make the predictive modeling process more accurate and smoother by prepping (estimates the quantities for 'train_scaled') and baking (assigns prepped data to an object) the data. The aforementioned top 5 predictors were used in this 'rec3'. Moreover, the categorical variables were factored and features with no variance were removed before this data was baked. Then, this processed data is assigned to 'folds3' and resampled on 5 subsets of the training set. Using 'fold3', we tune the hyperparameters to make the modeling process more efficient and perform grid specification to determine which hyperparameter values have high prediction accuracy (i.e. low prediction error). Next, we isolated the best hyperparameter values, so they can be used for the final boosting model. Lastly, we fitted the model and evaluated its prediction performance. Its RMSE score was 0.23, indicating that XG Boost is an adequate model. This website was used: [RBloggers](https://www.r-bloggers.com/2020/05/using-xgboost-with-tidymodels/)

@@

As seen on the table and visual, Random Forest has the highest RMSE and lowest R-squared estimate. LR has the second highest RMSE with a moderate R-squared estimate. And XG Boost has the lowest RMSE of 0.23 and the highest R-squared estimate of 0.99. The kNN model has the both the second lowest RMSE and second highest R-squared estimate. Since we are choosing the best model by the best RMSE, which is the lowest one, the XG Boost model is thus the best and final model based on our procedure. After fitting the XG Boost model to the testing data, we find that its RMSE and R-squared is 1.38 and 0.68, respectively. Since we are trying to predict annual average ambient air pollution of a given monitor, then these statistics mean that the amount of error in our model's predictions is about 1.38 $\mu$g/m$^3$, and that the features we used in our model are able to explain 68% of the variance in the data, which means this model does not fit/predict the data well.

@@

The model with only 'CMAQ' and 'aod' as features has an RMSE of 1.99. Since it is greater than 1, this means that 'CMAQ' and 'aod' are not good predictors for annual ambient air pollution in a given location. The RMSE of our model on the test data is 1.38. The RMSE of our model without 'CMAQ' and 'aod' is 1.73. This means that the model without 'CMAQ' and 'aod' has the better prediction performance since its RMSE is lower than our original model.

@@

```{r Q3-2}
set.seed(123)
# XGB model w/o cmaq and aod
# Preprocessing recipe
rec8 <- 
  recipes::recipe(value ~ lat+county_area+lon, data = training(dat_split)) %>%
  # Convert categorical variables to factors
  recipes::step_string2factor(all_nominal()) %>%
  # Combine low frequency factor levels
  recipes::step_other(all_nominal(), threshold = 0.01) %>%
  # Remove no variance predictors which provide no predictive information 
  recipes::step_nzv(all_nominal()) %>%
  prep()

@@

# Eval. model performance on train data
train_processed = bake(rec8, new_data=training(dat_split))
test_processed <- bake(rec8, new_data = testing(dat_split))

@@

**4. The dataset here did not include data from Alaska or Hawaii. Do you think your model will perform well or not in those two states? Explain your reasoning.**

Given that the top 5 features were 'CMAQ', 'lat', 'county_area', 'lon', and 'aod'. We think that XG Boost model will not perform well since our model did not perform well on the test dataset. It seems like we overfit our model, so fitting it to data from Hawaii and Alaska, which we expect to be veyr different from our training dataset, will yield similar results to fitting our model to the test dataset.

@@

The most challenging aspect of this project was figuring out the process to determine which predictors were the best to use for our models. We ran into many debugging issues and Googled a lot to resolve our uncertainties. We learned that finding the best predictors for a model can be a lengthy process and that running a new model (e.g. preprocessing steps for XG Boost) often may require lots of research into what is needed for the model to run before it can actually run.

@@

The RMSE of our model was 1.38 on the testing dataset, which means it does not predict well. It did not perform as well as we originally expected, especially since we saw that the RMSE of our model was 0.23 on the training dataset. We believe the model did not perform well because we overfit the model to the training data. This means that the model gives inaccurate predictions and cannot perform well for all types of new data, which is why despite the low RMSE on the training data, the model has poor prediction performance on the testing data.
